Vagueness and \emph{as-if}-generalization from imprecise imitation in signaling games

### 2015-10-23 paper/version_04/title_page.tex
Title page for submission to BPoS.

\documentclass[fleqn,reqno,10pt]{article}

%========================================
% Packages
%========================================

\usepackage[]{../../helpers/mypackages}
%\usepackage[natbib=true,style=authoryear-comp,backend=bibtex,doi=false,url=false]{biblatex}
%\bibliography{MyRefGlobal}
\bibliography{../../helpers/MyRefGlobal}
\bibliography{paper} 
\usepackage{../../helpers/myenvironments}
\usepackage{../../helpers/mycommands}
\usepackage{todonotes}
\usepackage{subcaption}

\usepackage{a4wide}

%========================================
% Standard Layout
%========================================

% Itemize
\renewcommand{\labelitemi}{\large{$\mathbf{\cdot}$}}    % itemize symbols
\renewcommand{\labelitemii}{\large{$\mathbf{\cdot}$}}
\renewcommand{\labelitemiii}{\large{$\mathbf{\cdot}$}}
\renewcommand{\labelitemiv}{\large{$\mathbf{\cdot}$}}
% Description
\renewcommand{\descriptionlabel}[1]{\hspace\labelsep\textsc{#1}}

% Figure Captions
\usepackage{caption} % use corresponding myfiguresize!
\setlength{\captionmargin}{20pt}
\renewcommand{\captionfont}{\small}
\setlength{\belowcaptionskip}{7pt} % standard is 0pt

%========================================
% Additional layout & commands
%========================================

\renewcommand{\Smixed}{\ensuremath{\mathrm{\mathbf{s}}}}
\renewcommand{\Rmixed}{\ensuremath{\mathrm{\mathbf{r}}}}

% Annotations
\newcommand{\mytodo}[2]{\todo[inline,color=yellow,author=#1]{#2}}
\newcommand{\question}[2]{\todo[inline,color=blue,author=#1]{#2}}
\newcommand{\answer}[2]{\todo[inline,color=green,author=#1]{#2}}

\newcommand{\rd}{\acro{rd}} % replicator dynamic
\newcommand{\rmd}{\acro{rmd}} % replicator mutator dynamic
\newcommand{\rdd}{\acro{rdd}} % replicator diffusion dynamic
\newcommand{\RD}{\ensuremath{\mathrm{RD}}} % replicator dynamic
\newcommand{\RDD}{\ensuremath{\mathrm{RDD}}} % replicator diffusion dynamic
\newcommand{\RMD}{\ensuremath{\mathrm{RMD}}} % replicator mutator
\newcommand{\icid}{\acro{icid}} % imprecise conditional imitation dynamic
\newcommand{\ICID}{\ensuremath{\mathrm{ICID}}} % imprecise conditional imitation dynamic

\newcommand{\Diff}{\ensuremath{\mathrm{D}}} % Difusion 
\newcommand{\Mutate}{\ensuremath{\mathrm{M}}} % Mutation

\newcommand{\imprecision}{\ensuremath{\alpha}} % imprecision
\newcommand{\toler}{\ensuremath{\beta}} % tolerance
\newcommand{\ns}{\ensuremath{n_s}} % number of states

\newcommand{\similarity}{\ensuremath{\mathrm{Sim}}} % similarity function

\newcommand{\myred}[1]{\textcolor{red}{#1}} 
\renewcommand{\Pr}{\ensuremath{P}}

% \doublespacing

\title{Vagueness and \emph{as-if}-generalization from imprecise imitation in signaling games}
% \author{Michael Franke \and Jos\'e Pedro Correia}
\date{}

@@

\section{Introduction}
\label{sec:introduction}

Many concepts and expressions are vague. A vague category knows clear
cases that fall under it, clear cases that do not, and also so-called
borderline cases. Borderline cases do not clearly apply, nor do they
clearly not apply, and there may be differences between borderline
cases in terms of how well they represent the category in
question. Vagueness does not seem to dramatically affect the success
of everyday communication, but it is troublesome for some of the most
prominent theories of language and meaning. This is especially so for the logico-positivist tradition of Frege, Russell and the young
Wittgenstein which is challenged by the paradoxes vagueness gives rise
to.

But there are other intriguing aspects about vagueness. The puzzle
that we are concerned with here is how vagueness could arise and be
maintained in the first place. This is a serious issue for
functionalist accounts that maintain that concepts and linguistic
meanings evolved driven towards efficiency. Since the existence of
unclear borderline cases seems to entail inefficiency of
categorization or communication, the challenge, succinctly put forward
by \citet{Lipman2009:Why-is-Language}, is to explain how vagueness can
persist under evolutionary pressure to be optimally discriminative. We
address \emph{Lipman's problem} as a technical challenge, namely to
find a conceptually sound and mathematically coherent formal model of
the evolution of categorization and language use that yields
by-and-large efficient categorization \emph{and} vagueness.

A number of authors have recently tried to explain why vagueness evolved as something that is
itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of limitations in information
processing \citep[e.g.][]{FrankeJager2010:Vagueness-Signa} or the result of particular
low-level learning strategies \citep[e.g.][]{OConnor2013:The-Evolution-o}. This paper
contributes to the latter line of thought. Concretely, we introduce a dynamic that describes
changes in aggregate population behavior as the result of imitation with probabilistic
confusability of similar stimuli. We show that stochastic noise can not only lead to vague, yet
relatively efficient signal use, but can also unify evolutionary outcomes and help avoid
suboptimal categorization: regularities in confusability of stimuli also regularize signaling
behavior and evolving categories. The system gives rise to generalization behavior without
agents actually generalizing.

The next section introduces the background against which the work presented here can be
appreciated. Section~\ref{sec:repl-diff-dynam} introduces a generalization of the replicator
dynamic which is derived from the idea that agents imitate other agents' behavior while
possibly confusing similar states. This imprecise imitation dynamic is explored in
Section~\ref{sec:exploring-rdd}. Section~\ref{sec:discussion} reflects and compares our
approach to others. Appendix~\ref{sec:formal-stuff} provides formal detail.

\section{Background}
\label{sec:background}

% The view that vagueness is a natural concomitant of cognitive
% limitations of language users has been formalized in a number of ways,
% using evolutionary game theory and certain generalizations of Lewis'
% signaling games \citep{Lewis_1969:Convention}, so called
% \emph{similarity-maximizing games}, or \emph{sim-max games}, for short
% \citep{Jager2007:The-Evolution-o,JagerRooijvan-Rooij2007:Language-Struct}.
% %\citep{FrankeJager2010:Vagueness-Signa,OConnor2013:Evolving-Percep,OConnor2013:The-Evolution-o}.
% Our contribution is best seen in relation to these accounts, as it also
% relies on sim-max games. Let's introduce these first, and then zoom in
% on the problem of vagueness.

\subsection{Sim-max games \& conceptual spaces}

Signaling games, as introduced by \citet{Lewis_1969:Convention}, have
a sender and a receiver. The sender knows the true state of the world,
but the receiver does not. The sender can select a signal, or message,
to reveal to the receiver, who then chooses an act. In Lewis' games,
if the receiver chooses the act that corresponds to the actual state,
the play is a success, otherwise a failure. Certain regular
combinations of sender signaling and receiver reaction make messages
meaningful, in the sense that their use is correlated systematically
to certain states or acts. To investigate the conditions under which
such meaning-generating behavior can evolve is a highly interesting
topic that we are only beginning to fully understand
\citep[e.g.][]{Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,Huttegger2007:Evolution-and-t,Pawlowitsch2008:Why-Evolution-d,Barrett2009:The-Evolution-o,HutteggerSkyrms2010:Evolutionary-Dy,Skyrms2010:Signals}.

Similarity-maximizing (short: sim-max) games are a variation of
Lewis' games where the receiver's actions are equated with the state
space (one can think of the actions as choosing states) and different states are allowed to be more or less
similar to one another. While Lewis' games treated communicative
success as a matter of black and white, sim-max games allow for a
gradient notion: the more similar the receiver's interpretation is to
the actual state, the better. Signaling games with utility-relevant
similarities in the state space are fairly standard in economics
\citep[e.g.][]{Spence1973:Job-market-sign,CrawfordSobel1982:Strategic-Infor},
but have received particular attention in a more philosophical context
for reasons that will become clear presently.

Formally, a sim-max game consists of a set of states $\States$, a set of messages $\Messgs$ typically with much
fewer messages than states, a probability distribution $\Pr \in \Delta(\States)$ such that
$\Pr(\state)$ gives the probability that state $\state$ occurs, a similarity metric\footnote{A
  metric is a function of distance between any two points in a given space. It should satisfy
  certain axioms that ensure behavior that one would intuitively expect from the words
  ``similarity'' and ``distance'' alone, but these details do not matter for the purposes of
  this paper.}
  % : similarity should always be non-negative, zero only for the distance
  % between an element and itself, symmetric, and satisfy the triangle inequality that
  % $\similarity(x,z) \le \similarity(x,y) + \similarity(y,z)$
on states $\similarity \mycolon \States \times \States \rightarrow \mathds{R}$ such that
$\similarity(\mystate{1}, \mystate{2})$ is the (perceptual) similarity between $\mystate{1}$
and $\mystate{2}$, and a utility function
$\Utils \mycolon \States \times \States \rightarrow \mathds{R}$ such that
$\Utils(\mystate{1}, \mystate{2})$ is the payoff for sender and receiver for a play with actual
state $\mystate{1}$ and receiver interpretation $\mystate{2}$. We identify the receiver's acts
with the states of the world, so that the game is one of guessing the actual state, so to
speak. The utility function should be a monotonically increasing function of similarity.
Non-probabilistic sender behavior can be represented by a pure strategy
$\Spure \in \Messgs^\States$, which deterministically defines which message would be used for
each state. Similarly, a pure receiver strategy is a function $\Rpure \in \States^\Messgs$.

\citet{JagerMetzger2011:Voronoi-Languag} showed that the evolutionarily stable states of
sim-max games are remarkably systematic: the evolutionarily stable states are demonstrably
so-called Voronoi languages.\footnote{Their results were obtained for games with infinitely
  many states in $n$-dimensional Euclidean space $\States \subseteq \mathds{R}^n$ and a
  quadratic loss function for utilities
  $\Utils(\mystate{1}, \mystate{2}) = - (\mystate{1} - \mystate{2})^2$.} Roughly put, a Voronoi
language is a pair of sender and receiver strategies, such that the sender strategy partitions
the state space into convex categories, while the receiver's interpretations are the central
spots in each category. A subset $X$ of $\mathds{R}^n$ is convex if, informally put, all points
in $X$ are connected via a straight line that lies entirely in $X$; $X$ has no gaps or
dents. For example, if \States is the unit interval and all states are equiprobable, a Voronoi
language with two messages could have the sender use one message exclusively for all points in
the lower half of the unit interval and another for all points in the upper half; the
receiver's interpretations of messages are the central points, .25 and .75, in the respective
intervals.  See Figure~\ref{fig:Voronoi-language-example} for an illustration.

\begin{figure}
  \centering

  \begin{tikzpicture}[scale=3]

        \begin{scope}

          \draw[-,thick] (0,0) -- (1.5,0);

          \draw (0,0) node {\color{red}{\big{[}}};

          \draw (.75,0) node {\color{red}{\big{)}}};

          \draw (.76,0) node {\color{red}{\big{(}}};

          \draw (1.5,0) node {\color{red}{\big{]}}};

            \draw (.325,.15) node
            {\color{red}{$\Spure(\mymessg{1}  \probbar \cdot)$}};

            \draw (1.075,.15) node
            {\color{red}{$\Spure(\mymessg{2} \probbar \cdot)$}};

            \node at (.375,0) (node1) {\color{blue}{$\bullet$}};

            \node at (.375,-.3) (label1)
            {\color{blue}{${\Rpure}(\cdot \probbar \mymessg{1})$}};

            \draw[-,color=blue] (node1) -- (label1);

            \node at (1.125,0) (node2) {\color{blue}{$\bullet$}};

            \node at (1.125,-.3) (label2)
            {\color{blue}{${\Rpure}(\cdot \probbar \mymessg{2})$}};

            \draw[-] (node2) -- (label2);

            \draw (0,-.2) node {$0$};

            \draw (.75,-.2) node {$0.5$};

            \draw (1.5,-.2) node {$1$};

          \end{scope}

        \end{tikzpicture}

        \caption{Example of a Voronoi language on $\States = [0,1]$. The pure sender strategy
          $\Spure$ uses one signal for the lower half, and another for the upper half of the
          unit interval. The pure receiver strategy $\Rpure$ selects the central elements in
          the respective intervals. See \citet{JagerMetzger2011:Voronoi-Languag} for further
          details.}
  \label{fig:Voronoi-language-example}
\end{figure}

This result is interesting because it demonstrates that signaling can
impose orderly categories on a metric space, without that being the
ulterior purpose of it all. Sender and receiver can be distinct
entities, whose purpose is to communicate effectively about the actual
state. In that case, evolving Voronoi languages would explain why
\emph{linguistic categories} are well-behaved and orderly in the way
they appear to be. More abstractly, sender and receiver can also be
thought of as distinct modules in a single system, where the first
module must discretize the information it is fed by selecting a small
sample of, suggestively, category labels. These are passed to a second
module that tries to decode the original information. In this case,
evolving Voronoi languages would explain why \emph{conceptual
  categories} are well-behaved and orderly in the way that they appear
to be.

Seen in this light, sim-max games may provide a foundation to approaches in cognitive semantics
that rely on the notion of conceptual spaces.
\citet[][70--77]{Gardenfors2000:Conceptual-Spac}, for example, has prominently argued that
natural categories are convex regions in conceptual space. If the conceptual space has a
suitable metric, convex categories can be derived from a set of prototypes. The category
corresponding to prototype $p$ is the set of points that are more similar to $p$ than to any
other. In this way, \citeauthor{Gardenfors2000:Conceptual-Spac} argues, an efficient categorization
system can be obtained: storing the prototypes lets us recover the categories without having to
store each category's extension. However, what is left unexplained is where the prototypes
come from, and why we would not see just any distribution of prototypes as an equally efficient
classification system. This is where sim-max games can contribute a principled approach to
deriving, in an independent way, not only convex categories but also prototypical exemplars
belonging to them.  These ideas, and more, are developed further by
\citet{Jager2007:The-Evolution-o,JagerRooijvan-Rooij2007:Language-Struct,JagerMetzger2011:Voronoi-Languag,OConnor2014-OCOEPC},
among others.

\subsection{Vague signaling in sim-max games}

This outline of an approach to categorization using sim-max games
leaves many problems unaddressed. One of them is that natural
categories for continuously variable stimuli usually do not have
unique, point-valued prototypes and clear category boundaries. We
would like to account for the possibility of such vagueness, in
particular: (i) clear positive examples of a vague category should
show a gradient transition to clear negative examples;
(ii) prototypes should likewise be gradient regions, peaking at the
center of the vague category they represent.

\citet{DouvenDecock2011:Vagueness:-A-Co} show that
\citeauthor{Gardenfors2000:Conceptual-Spac}'s conceptual spaces
approach can be extended to account for the existence of borderline
cases. From the assumption that prototypes are extended, yet convex
regions in conceptual space, a construction algorithm is available
that yields ``collated Voronoi diagrams'' with thick boundaries
representing borderline
regions. \citet{DecockDouven2012:What-is-Graded-} show further how it
is possible to arrive at a gradient transition between categories, by
weighing in the distance of different borderline cases to various
prototypical regions. This accounts for the first of the two
desiderata mentioned above, but still assumes that crisp prototype
regions must be given.

Alternative approaches are taken by, for example, \citet{FrankeJager2010:Vagueness-Signa} and
\citet{OConnor2013:The-Evolution-o}, who show different ways how the above desiderata can be met by evolving
strategies in sim-max games. The approach we take here is similar in spirit, but different in
relevant detail. A more in-depth comparison is deferred until Section~\ref{sec:discussion}. Let
us first look into our own proposal in more detail.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Imprecise imitation}
\label{sec:repl-diff-dynam}

Signaling agents can adapt their dispositions to act, given some feedback about their past
success, in multiple ways. Usually, we would assume that changes in behavior should, at least
on average, lean towards increasing chances of communicative success. Such behavioral
adaptations can be described at different levels of abstraction. At the level of individual
agents, we can picture a more or less idealized process of how each agent adapts dispositions
for future actions based on various pieces of information available to the agent. More
abstractly, at the level of a population of agents, we can describe how average behavioral
dispositions will evolve. The population-level perspective abstracts over small stochastic
fluctuations and zooms in on the general tendency or direction of evolution that ensues from
behavior at the agent-level.

To better understand the interaction of confusability of stimuli and selective pressure towards
successful communication, we look at a population-level dynamic that describes the most likely
evolutionary path of a population of signaling agents who imitate other agents' behavior but
are liable to confuse states for one another. In the special limiting case where state
confusability vanishes, the process is just the well-known replicator dynamic
\citep{TaylorJonker1978:Evolutionary-St}, which should therefore be briefly reviewed first.

\subsection{Replicator dynamic in behavioral strategies}
\label{sec:repl-dynam-behav}

Fix a sim-max game with finite states $\States$ and messages $\Messgs$. As usual, we assume
that the receiver chooses states in $\States$ in response to messages. Let
$\Pr(\cdot) \in \Delta(\States)$ be the prior distribution over states and
$\Util \mycolon \States \times \States \rightarrow \mathds{R}$ the utility function shared by
senders and receivers in the population. A \emph{behavioral strategy} is a function that maps
an agent's choice points to a probability distribution over available
choices.\footnote{\label{fn:mixedStrats}Our focus is on behavioral strategies, not \emph{mixed
    strategies}, i.e., probability distributions over functions from each choice point to an
  act, such as $s \in \Delta(\Messgs^{\States})$.  Dynamics on behavioral strategies assume
  that agents can adapt their behavior locally, i.e., independently at each choice point. Our
  focus on behavioral strategies greatly reduces the complexity of the dynamic and simplifies
  numerical simulations. But it also seems the more plausible choice for imitation-based update
  protocols of the kind we consider here: agents only observe how, on some occasion, some other
  agent behaved in \emph{one} particular situation, not how that agent would behave in
  \emph{all} relevant choice situations; they imitate the use of a single word, so to speak,
  not a whole lexicon. (See \citet{Cressman2003:Evolutionary-Dy} for more on the difference
  between dynamics on mixed or behavioral strategies.)} The sender's behavioral strategies are
functions $\Sstrat \in \Delta(\Messgs)^\States$, thus mapping each state $\state \in \States$
to a probability of each message $\messg \in \Messgs$ being sent in $\state$; the receiver's
are functions $\Rstrat \in \Delta(\States)^\Messgs$, thus mapping each message
$\messg \in \Messgs$ to a probability of each interpretation $\state \in \States$ being chosen
in response to $\messg$. Although behavioral strategies are probabilistic, evolutionary
modeling usually imagines that every individual agent has a non-probabilistic
strategy. Behavioral strategies then capture average population behavior. Assuming a virtually
infinite population, the number $\Sstrat(\messg \probbar \state)$, for instance, is then the
probability that a randomly sampled sender would send message $\messg$ if the actual state was
$\state$. Similarly, $\Rstrat(\state \probbar \messg)$ is then the probability with which a
randomly sampled receiver interprets $\messg$ as $\state$. A play of single evolutionary game
with behavioral strategies is illustrated in Figure~\ref{fig:sim-max-illustration}.

\begin{figure}
  \centering

  \begin{tikzpicture}[->,auto,semithick,>=stealth',every label/.style={gray},node distance=0.9cm]

  \node (1) {};
  \node (2) [right=of 1,label=below:$\Pr(\mystate{1})$] {$\mystate{1} \in \States$};
  \node (3) [right=of 2,label=below:$\Sstrat(\messg \probbar \mystate{1})$] {$\messg \in \Messgs$};
  \node (4) [right=of 3,label=below:$\Rstrat(\mystate{2} \probbar \messg)$] {$\mystate{2} \in \States$};

  \path
    (1) edge node {N} (2)
    (2) edge node {S} (3)
    (3) edge node {R} (4);

  \end{tikzpicture}

  \caption{A round of play in a sim-max game with behavioral strategies: Nature (N) chooses a
    state $\mystate{1}$ with probability $\Pr(\mystate{1})$; a random sender (S) selects a
    message $\messg$ with probability $\Sstrat(\messg \probbar \mystate{1})$; a random receiver
    (R) selects a state $\mystate{2}$ with probability $\Rstrat(\mystate{2} \probbar \messg)$.
    Payoff for both sender and receiver is given by $\Utils(\mystate{1}, \mystate{2})$.}
   %which is typically proportional to state similarity $\similarity(\mystate{1}, \mystate{2})$.}
  \label{fig:sim-max-illustration}
\end{figure}

The \emph{expected utility} of choices at each choice point is:\footnote{The notation
  $\propto$, for ``proportional to'', that is used here and hereafter means that the right-hand
  side might still need to be normalized so as to have probabilities sum to one. In general,
  writing $P(x) \propto f(x)$ for any function $f \mycolon X \rightarrow \mathds{R}$ is
  shorthand for $P(x) = \frac{f(x)}{\sum_{x' \in X} f(x')}$. }
\begin{align*}
  \EU(\messg, \state, \Rstrat) & = \sum_{\state \in \States}
  \Rstrat(\state' \probbar \messg) \myts U(\state, \state') \\
  \EU(\state', \messg, \Sstrat) & = \sum_{\state \in
    \States} P(\state \probbar \messg)  \myts
  U(\state, \state')\,, \text{\ \  where } &
P(\state \probbar \messg) & \propto \Pr(\state) \myts \Sstrat(\messg \probbar \state) \,.
\end{align*}
Expected utilities are sums, over all possible concrete outcomes, of the utilities of these
outcomes, weighted by how likely they occur. For instance, the expected utility for the sender
of choosing message $\messg$, given state $\state$ and a receiver strategy $\Rstrat$ is the
sum, for each interpretation $\state'$, of the probability that some random receiver will
choose $\state'$ given $\messg$ times the actual utility of triple $\state$, $\messg$ and
$\state'$.

The discrete-time replicator dynamic tracks changes in frequency of choices in the population
as proportional to their expected utilities
\citep[e.g.][]{HofbauerSigmund1998:Evolutionary-Ga}:\footnote{The formulation given in
  Equation~(\ref{eq:RDDiscrete}) is adequate only for cases like the one we will be looking at
  in Section~\ref{sec:exploring-rdd}, where utilities are always non-negative and expected
  utilities are always positive.}
\begin{align}
  \label{eq:RDDiscrete}
  \Sstrat'(\messg \probbar \state) & \propto \Sstrat(\messg \probbar \state) \myts
    \EU(\messg, \state,\Rstrat)\,, & \Rstrat'(\state \probbar \messg) & \propto \Rstrat(\state \probbar \messg) \myts
    \EU(\state, \messg,\Sstrat)\,.   
\end{align}
For a given choice point, say a state $\state$, the probability of seeing $\messg$ played by an
average agent in the population \emph{after} the update is proportional to the probability of
seeing it \emph{before} the update, which is $\Sstrat(\messg \probbar \state)$, times the
expected utility of $\messg$ at state $\state$. Intuitively put, frequencies of choices change
by a gradient of current frequency and a measure of how good they are.

The replicator dynamic is an abstract population-level dynamic that describes the mean expected
change of behavioral dispositions in a population of signalers. There are several ways of
deriving the replicator dynamic from agent-level processes of behavioral adaptation. We focus
here on one of the simplest: \emph{imitation of success}
\citep[see][]{Sandholm2010:Population-Game}. The intuitive idea is the following (see
Appendix~\ref{sec:cond-imit-succ} for a derivation of the standard replicator dynamic from this
update scheme).  Every now and then a random agent gets a chance to alter his behavior for one
of his choice points. Say, this is a sender who gets to ``reconsider'' his choice of message
for state $\state$. The revising agent then observes what some random agent does at $\state$,
say $\messg$, and will henceforth play $\messg$ in $\state$ with a probability given by the
expected utility of $\messg$ for state $\state$.

\subsection{Noisy replicator dynamic}
\label{sec:noisy-repl-dynam}

Imitation of success, as described above, presupposes that agents make no mistakes when
observing states, or choosing interpretations. This may not always be an appropriate
assumption, especially when some states can be perceptually similar and therefore likely to be
confused for one another. Therefore, imitation dynamics could be affected by at least two
sources of probabilistic noise:

\begin{enumerate}
\item \emph{observation noise:} whenever a state $\mystate{a}$ actually occurs, the probability
  that an agent observes it as $\mystate{o}$ is $P_o(\mystate{o} \probbar \mystate{a})$;
\item \emph{realization noise:} whenever an agent intends to realize interpretation
  $\mystate{i}$, the probability that $\mystate{r}$ is realized is
  $P_r(\mystate{r} \probbar \mystate{o})$.
\end{enumerate}

\noindent A round of play of a sim-max game with these two sources of confusability of states
is pictured in Figure~\ref{fig:imprecise-sim-max-illustration}. Note that, and this is crucial,
senders respond to observations not actual states, receiver strategies determine intentions not
realized states, but payoff is calculated based on actual and realized states. Behavioral
strategies thus encode what agents actually do from their subjective point of view; or, put
differently, what they would do in a noise-free world. The noise-free situation depicted in
Figure~\ref{fig:sim-max-illustration} is then the special case where
$P_o(\mystate{o} \probbar \mystate{a}) = 1$ iff $\mystate{a} = \mystate{o}$ and
$P_r(\mystate{r} \probbar \mystate{i}) = 1$ iff $\mystate{i} = \mystate{r}$.

\begin{figure}
  \centering

  \begin{tikzpicture}[->,auto,semithick,>=stealth',every label/.style={gray},node distance=0.9cm]

  \node (1) {};
  \node (2) [right=of 1,label=below:$P_a(\mystate{a})$] {$\mystate{a} \in \States$};
  \node (2') [right=of 2,label=below:$P_o(\mystate{o} \probbar \mystate{a})$] {$\mystate{o} \in \States$};
  \node (3) [right=of 2',label=below:$\Sstrat(\messg \probbar \mystate{o})$] {$\messg \in \Messgs$};
  \node (4) [right=of 3,label=below:$\Rstrat(\mystate{i} \probbar \messg)$] {$\mystate{i} \in \States$};
  \node (4') [right=of 4,label=below:$P_r(\mystate{r} \probbar \mystate{i})$] {$\mystate{r} \in \States$};

  \path
    (1) edge node {N} (2)
    (2) edge node {} (2')
    (2') edge node {S} (3)
    (3) edge node {R} (4)
    (4) edge node {} (4');

  \end{tikzpicture}

  \caption{A round of play in a sim-max game with probabilistic confusability of states: Nature
    (N) chooses an actual state with probability $P_a(\mystate{a})$; a randomly sampled sender
    (S) observes $\mystate{o}$ with probability $P_o(\mystate{o} \probbar \mystate{a})$ and
    subsequently selects $\messg$ with probability $\Sstrat(\messg \probbar \mystate{o})$; a
    randomly sampled receiver (R) intends to realize interpretation $\mystate{i}$ with
    probability $\Rstrat(\mystate{i} \probbar \messg)$ but actually realizes interpretation
    $\mystate{r}$ with probability $P_r(\mystate{r} \probbar \mystate{i})$. Payoff for both
    sender and receiver is given by $\Utils(\mystate{a}, \mystate{r})$.}
  %which is typically proportional to state similarity $\similarity(\mystate{a}, \mystate{r})$.}
  \label{fig:imprecise-sim-max-illustration}
\end{figure}

The presence of observation and realization noise also affects imitation of successes. Here, we
focus on the main ideas. Formal detail is provided by Appendix~\ref{sec:imit-succ-with}. Take a
sender who gets to revise behavior for state $\state$. What agents can plausibly revise by
imitation is their pure strategy, which maps perceived states onto messages. When an agent gets
to revise his strategy for the perceived state $\state$, this need not necessarily be the
actual state. Moreover, when that agent observes $\state$, another agent may perceive yet a
different state. Given what that latter agent perceives, his actual (pure) strategy will
determine what he plays. In sum, to describe how likely the potential imitator observes a
message choice $\messg$, we are interested in the conditional probability
$P_o(\messg \probbar \state)$ that some other random agent selects $\messg$ when the first
agent perceives $\state$. This $P_o(\messg \probbar \state)$ is derived from the prior
probability of states, the current sender population behavior $\Sstrat$ and the given
observation noise (see Appendix~\ref{sec:imit-succ-with}). Eventually, the imitating agent adopts
$\messg$ as his choice for $\state$ with probability given by the expected utility of sending
$\messg$ when perceiving state $\state$. Expected utility should, of course, take the
probabilistic confusability of states into account as well.

Similar considerations apply to the receiver side. If an agent gets a chance to change his
intended interpretation of $\messg$, we need to look at the conditional probability
$P_o(\state \probbar \messg)$ of observing another agent realize interpretation $\state$ given
that the first agent (and therefore the second as well) perceived $\messg$. This depends on
observation and realization noise, as well as on the current receiver population behavior
$\Rstrat$.

Appendix~\ref{sec:imit-succ-with} shows how imprecise imitation of this sort leads to mean changes in population
frequencies of choices that can be covered by the following discrete-time formulation:
\begin{align}
  \label{eq:nRD}
  \Sstrat'(\messg \probbar \state) & \propto P_o(\messg \probbar \state) \myts \EU(\messg,
  \state, \Rstrat)\,, & \Rstrat'(\state \probbar \messg) & \propto P_o(\state \probbar \messg)
  \myts \EU(\state, \messg, \Sstrat)\,.
\end{align}
\noindent This looks very much like the discrete-time formulation of the standard replicator
dynamic in (\ref{eq:RDDiscrete}), but there are, of course, the aforementioned
differences. Firstly, expected utility here takes stochastic confusability of states into
account. Secondly, where the standard replicator dynamic had probabilities
$\Sstrat(\messg \probbar \state)$ and $\Rstrat(\state \probbar \messg)$, we now have
$P_o(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ respectively. If there is no
observation or realization noise, $P_o(\messg \probbar \state)$ reduces to
$\sigma(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ reduces to
$\Rstrat(\state \probbar \messg)$. The noisy replicator dynamic in (\ref{eq:nRD})
conservatively extends the classic case in (\ref{eq:RDDiscrete}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Exploring imprecise imitation}
\label{sec:exploring-rdd}

How does a tendency to confuse similar states interact with selective pressure towards more
efficient signaling strategies under the imprecise conditional imitation dynamic? If confusion
probabilities are moderate and regular, in that they track similarity of states, we might
expect a regularizing effect also on evolving signaling strategies. Indeed, we hypothesize that
confusion of states can give rise to population-level aggregate behavior that looks \emph{as
  if} information about what is good for one state percolates also to similar states. In other
words, imprecise imitation may make signaling behavior on average look like agents generalize
across similar states, even if no agent actually generalizes. To explore whether confusion of
states can have this effect, we turn to numerical simulation.

\subsection{Setting the stage}
\label{sec:setting-stage}

To obtain concrete results, we must fix how to represent states, similarity of states, the
conditional confusion probabilities $P_o$ and $P_r$ and the utilities of our sim-max
games. Confusion probabilities between states should be a function of perceptual similarity:
the more similar two states are, the more likely they could be mistaken for each other.

Let the state space consist of $n \ge 2$ states that are equally spaced across the unit
interval, including $0$ and $1$. All states occur, for simplicity, with the same probability
(i.e., $P_a$ is uniform).  The distance $\card{\mystate{i} - \mystate{j}}$ is the objective,
physical similarity between two states $\mystate{i}$ and $\mystate{j}$. Distance in physical
space feeds into a perceptual similarity function, as described by
\citet{Nosofsky1986:Attention-Simil}:
\begin{align*}
  \similarity(\mystate{i}, \mystate{j} \myts ; \myts \imprecision) =
      \begin{cases}
    1 & \textrm{if } \imprecision = 0 \textrm{ and } \mystate{i} = \mystate{j} \\
    0 & \textrm{if } \imprecision = 0 \textrm{ and } \mystate{i} \neq \mystate{j} \\
 \expo \left ( -  \frac{\card{\mystate{i} - \mystate{j}}^2}{ \imprecision^2} \right ) & \textrm{otherwise,} \\
    \end{cases}
\end{align*}
where $\imprecision \ge 0$ is an imprecision parameter. When $\imprecision=0$ agents perfectly
discriminate between states; when $\imprecision \rightarrow \infty$ agents cannot discriminate
states at all. Figure~\ref{fig:NosofskySim} gives an impression of Nosofsky-similarity for
different parameter values. Other formalizations of perceptual similarity are possible,
including ones that allow for different discriminability in different areas of the state space,
but we stick with Nosofsky's similarity function for the time being, because it is
mathematically simple, yet an established notion in mathematical psychology.

\begin{figure}
  \centering

  \includegraphics[width=0.7\textwidth]{plots/NosofskySim.pdf}

  \caption{Examples of Nosofsky-similarity for different values of
    imprecision.}
  \label{fig:NosofskySim}
\end{figure}

We further assume that the probability of confusing any two states $\mystate{i}$ and
$\mystate{j}$ is proportional to their perceived similarity and, to keep matters simple, that
observation noise $P_o$ is simply the same as realizational confusability $P_r$ and
that both are governed by the same imprecision parameter $\alpha$:
\begin{align*}
  P_o(\mystate{o} \probbar \mystate{a}) & \propto \similarity(\mystate{o}, \mystate{a} \myts ; \myts
  \imprecision)\,, &   P_r(\mystate{r} \probbar \mystate{i}) & \propto \similarity(\mystate{r}, \mystate{i} \myts ; \myts
  \imprecision) \,.
\end{align*}
For $\imprecision = 0$, we obtain trivial confusion probabilities: everything is reduced to
perfect imitation and the replicator dynamic. For $\imprecision > 0$, any state can be confused
for any other state with some positive probability. For the uninteresting case of
$\imprecision \rightarrow \infty$, confusion is maximal and every state can be perceived or
realized as any other state with probability $\nicefrac{1}{\card{\States}}$.

As for utility, we define it in terms of similarity and introduce another free parameter, $\toler \ge 0$.
 The intention is for this parameter to model the amount of tolerable pragmatic slack, which we felt should be allowed to vary separately from perceptual imprecision:
\begin{align*}
  \util(\mystate{i}, \mystate{j} \myts ; \myts \toler) =
      \similarity(\mystate{i}, \mystate{j} \myts ; \myts \toler)\,.
\end{align*}
This choice of utility function is governed partly by convenience, but also because we believe it has the right general properties for a communicative payoff function.
Unlike utilities that are, say, linearly or
quadratically decreasing in physical distance
\citep[c.f.][]{JagerMetzger2011:Voronoi-Languag,FrankeJager2010:Vagueness-Signa}, utilities
that are exponentially decreasing in negative quadratic distance can model situations where a
small amount of imprecision in communication is tolerable, whereas similarly small differences
in intolerably far away interpretations matter very little, with a smooth transition between
these regimes \citep[c.f.][]{OConnor2013:The-Evolution-o}.

In sum, to structure thinking about the behavior of our imprecise imitation dynamic, the system
is governed by three parameters: the number $\ns = \card{T}s$ of states in the state space,
imprecision $\imprecision$ and tolerance $\toler$.

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}

\title{Vagueness and \emph{as-if}-generalization from imprecise imitation in signaling games}

\author{
Michael Franke%
\thanks{Department of General Linguistics, Faculty of Humanities, University of Tübingen}\\
\texttt{mchfranke@gmail.com}
\and
Jos\'e Pedro Correia\\
\texttt{zepedro.correia@gmail.com}
}

\date{}

@@
